---
title: 'Heart Attack Data from Kaggle'
author: 'Mark Newman'
date: '2021-04-13'
editor_options: 
  chunk_output_type: console
---

<!--
This RMD can and must be runnable in a stand alone manner.
However, it is intended to be called from other RMDs using the r chunk's `child` parameter.
-->

Download the data from [Kaggle](https://www.kaggle.com/nareshbhat/health-care-data-set-on-heart-attack-possibility).
It needs to be downloaded first because it is behind Kaggle's Sign In page.
**NOTE**: The file is in _UTF-8-BoM_.
We need to re-encode the file to regular _UTF-8_ before starting.

```{r label = 'load data'}
data <- read.csv('../data/heart.csv')
```

`factor()` the data as necessary based on the data dictionary.

```{r label = 'data dictionary conversion', echo = F}
data$sex <- factor(data$sex, levels = 0:1)
levels(data$sex) <- c('female', 'male')

data$cp <- factor(data$cp, levels = 0:3)
levels(data$cp) <- c('typical angina', 'atypical angina', 'non-anginal pain', 'asymptomatic')

data$fbs <- ordered(data$fbs, levels = 0:1)
levels(data$fbs) <- c('<= 120 mg/dl', '> 120 mg/dl')

data$restecg <- factor(data$restecg, levels = 0:2)
levels(data$restecg) <- c('normal', 'abnormality', 'hypertrophy')

data$exang <- factor(data$exang, levels = 0:1)
levels(data$exang) <- c('no', 'yes')

data$slope <- factor(data$slope, levels = 0:2)
levels(data$slope) <- c('upsloping', 'flat', 'downsloping')

data$thal <- factor(data$thal, levels = 1:3)
levels(data$thal) <- c('normal', 'fixed defect', 'reversable defect')

data$target <- ordered(data$target, levels = 0:1)
levels(data$target) <- c('less chance', 'more chance')
```

```{r label = 'qa factor known fail', echo = F}
t1 <- nrow(data)
data <- data[!is.na(data$thal),]
t2 <- t1 - nrow(data)
```

**NOTE**: The [Kaggle](https://www.kaggle.com/nareshbhat/health-care-data-set-on-heart-attack-possibility), [UCI MLR](https://archive.ics.uci.edu/ml/datasets/Heart+Disease), and [source paper]( https://doi.org/10.1378/chest.94.2.380) all represent three levels for `thal`.
It is therefore interesting that the dataset has a level `0`.
We will treat this as a `NA` and remove it.
This results in a loss of `r t2` observations (~`r round(100*t2/t1, 1)`%).

```{r label = 'qa factor known fail cleanup', echo = F}
rm(t1,t2)
```

QA check the conversion by testing for `NA`s in the factor variables.

```{r label = 'qa factor'}
for(col in colnames(data)) {
  t1 <- data[,col]
  if(is.factor(t1) & any(is.na(t1))) {
    warning(paste0("QA: Column '", col, "' has NAs"))
  }
  rm(t1)
}
rm(col)
```

Mark the `class` variable in both string and formula form.

```{r label = 'formula'}
class <- 'target'
formula <- target ~ .
```
